### 2025-07-17 manuscript/main.tex
test description
abtrsct abbrivs.

\documentclass[11pt]{report}
\usepackage[paper=a4paper,margin=2.5cm]{geometry}
\usepackage[outputdir=out, newfloat]{minted}
\usepackage{multicol}
\usepackage{tikz}
\usetikzlibrary{
  arrows.meta,      % nicer arrow tips
  shapes.geometric, % rectangle, diamond, etc.
  positioning,      % easy relative placement
  chains,           % flow‐style “chain” syntax
  calc             % coordinate arithmetic
}
\tikzset{
  process/.style = {
    rectangle,
    draw,
    minimum width=3cm,
    minimum height=1cm,
    text centered,
    on chain
  },
  arrow/.style   = {
    -Stealth,
    thick
  },
  start chain    = going below,  % default chain direction
  node distance  = 8mm           % space between chain nodes
}
\usepackage{amsfonts}
\usepackage{fontspec}

@@

\begin{abstract}
  % Your abstract text here.
  all slash code coantined text hs to be revisted due to the detokenize
\end{abstract}

@@

@@

@@

% Full Miniprob grammar here.

\chapter{Expanded Code Listings}
% Additional code excerpts.
\inputminted[linenos, breaklines]{yaml}{samples/mini-prob.langium}
\captionof{listing}{Langium grammar definition of MiniProb (LL(2) parsable)}
\label{list:langium-grammar}

>>>>

\documentclass[11pt]{report}
\usepackage[paper=a4paper,margin=2.5cm]{geometry}
\usepackage[outputdir=out, newfloat]{minted}
\setminted{
  fontsize=\small
}
\usepackage{multicol}
\usepackage{tikz}
\usetikzlibrary{
  arrows.meta,      % nicer arrow tips
  shapes.geometric, % rectangle, diamond, etc.
  positioning,      % easy relative placement
  chains,           % flow‐style “chain” syntax
  calc             % coordinate arithmetic
}
\tikzset{
  process/.style = {
    rectangle,
    draw,
    minimum width=3cm,
    minimum height=1cm,
    text centered,
    on chain
  },
  arrow/.style   = {
    -Stealth,
    thick
  },
  start chain    = going below,  % default chain direction
  node distance  = 8mm           % space between chain nodes
}
\usepackage{amsfonts}
\usepackage{fontspec}

@@

\begin{abstract}
  % Your abstract text here.
  ??all slash code coantined text hs to be revisted due to the detokenize\\

\noindent Cyber-Physical Systems demand rigorous, automated verification to ensure safety and correctness under uncertainty. This thesis presents the design and
implementation of a Visual Studio Code language extension for \textbf{MiniProb} (\ref{sec:miniprob}), a domain-specific probabilistic programming language tailored to formal model checking of
recursive, stochastic processes. Building on the Langium framework - written in TypeScript and running on Node.js - and the Language Server Protocol, a formal EBNF grammar for
MiniProb was defined and used within Langium to generate a parser that produces both Concrete Syntax Trees (CSTs) and typed Abstract Syntax Trees (ASTs).
Custom services - including semantic validation, type checking, cross-reference resolution, and some code completion features - were developed within Langium.

\noindent  The extension, packaged as a .vsix and available on GitHub, streamlines MiniProb development by lowering the barrier to entry for probabilistic model
authors and facilitating direct translation of MiniProb programs into probabilistic operator precedence automata for formal verification.

\noindent This work demonstrates that modern editor tooling, when coupled with robust language engineering frameworks, can significantly enhance the usability and
accessibility of specialized DSLs in safety-critical CPS contexts.

\end{abstract}

@@

conlsuio: mention no dode comülpeteion for structures only refs, not optimal and how it would be implemetned

@@

\chapter{Abbreviations}
\begin{description}
  \item[API] Application Programming Interface
  \item[AST] Abstract Syntax Tree
  \item[BNF] Backus-Naur Form
  \item[BOOL] Boolean
  \item[CAV] Computer-Aided Verification
  \item[CFG] Context-Free Grammar
  \item[CLI] Command-Line Interface
  \item[CPS] Cyber-Physical System
  \item[CS] Computer Science
  \item[CST] Concrete Syntax Tree
  \item[DFA] Deterministic Finite Automaton
  \item[DI] Dependency Injection
  \item[DSL] Domain-Specific Language
  \item[EBNF] Extended Backus-Naur Form
  \item[ESM] ECMAScript Module
  \item[JSON-RPC] JavaScript Object Notation -- Remote Procedure Call
  \item[LL] Left-to-right, Leftmost derivation
  \item[LR] Left-to-right, Rightmost derivation
  \item[LSP] Language Server Protocol
  \item[LTL] Linear Temporal Logic
  \item[MCMC] Markov Chain Monte Carlo
  \item[OPA] Operator Precedence Automaton
  \item[OPM] Operator Precedence Matrix
  \item[PDA] Pushdown Automaton
  \item[PEG] Parsing Expression Grammar
  \item[POTL] Procedence Operator Temporal Logic
  \item[PPL] Probabilistic Programming Language
  \item[SMC] Statistical Model Checking
  \item[SMT] Satisfiability Modulo Theories
  \item[URI] Uniform Resource Identifier
  \item[VI] Variational Inference
  \item[VS] Visual Studio
  \item[VSCE] Visual Studio Code Extension
  \item[WS] Web Services
  \item[XML] eXtensible Markup Language
\end{description}

@@

\chapter{Expanded Code Listings}
\inputminted[linenos, breaklines]{yaml}{samples/mini-prob.langium}
\captionof{listing}{Langium grammar definition of MiniProb (LL(2) parsable)}
\label{list:langium-grammar}
